Schema Change: Challenge Problems

### 2024-05-13 paper.tex
beef up Extract Entity

\usepackage[backend=biber]{biblatex}
%\usepackage{natbib}
\addbibresource{paper.bib}

@@

Some tools provide a Domain Specific Language (DSL) to describe schema evolution. Rails Migrations~\cite{RailsMigrations} embeds a DSL in Ruby to describe a range of common schema evolutions. However that falls short of the range of evolutions described by \citet{ambler06}, requiring the addition of custom Ruby or SQL code. \citet{curino08} spawned a stream of research on Database Evolution Languages (DEL) by defining a Schema Modification Operator (SMO) as ``a function that receives as input a relational schema and the underlying database, and produces as output a (modified) version of the input schema and a migrated version of the database''. SMOs can also rewrite queries to accomodate schema changes. \citet{herrmann15} defined a relationally complete DEL and then extended it into a Bidirectional Database Evolution Language (BiDEL)~\cite{herrmann17}. BiDEL appears capable of handling the basic requirements of our Extract Entity and Multiplicity Change challenges, though Split/Merge Entity is less clear. It provides schema divergence by supporting multiple schema within one database, but would need some extension to handle data divergence.

Schema evolution is also a problem for NoSQL~\cite{sadalage12} databases. While such databases are sometimes called ``schemaless'' in effect that means the schema is left implicit and tools must try to infer it~\cite{storl20, storl22}. \citet{Cambria} uses lenses~\cite{Foster2007} for bidirectional transformation of JSON. \citet{scherzinger13} define a set of operators like the relational SMOs discussed above. \citeauthor*{chillon21}~\cite{chillon21, chillon22} offer a more comprehensive set of SMOs that may be capable of handling Extract Entity and Multiplicity Change but not Split/Merge Entity nor divergence. None of the above approaches to NoSQL evolution have yet extended into updating code or rewriting queries like their SQL cousins.

\section{Extract Entity}

\subsection{Context}
In this challenge we consider a typical example of how the design of a database evolves as new needs are discovered. Much has been written on the subject of data modeling methodologies and the theory of relational database normalization~\cite{Molina08}. But that work often focuses on determining the correct data model up front before the system goes live, which would indeed be the optimal solution if only we were omniscient. The Extract Entity challenge confronts the practical reality of evolving a relational data model in flight.

\subsection{Example}
The Acme Corporation needs to record orders from various customers for various products. The simplest implementation is a spreadsheet with a row for each order and columns for information about the customer and product. It might look like this:

\begin{table}[h!]
\centering
  \begin{tabular}{ |l|c|c|l|l|}
   \hline
   Item & Quantity & Ship Date & Customer name & Customer address \\
   \hline \hline
   Anvil & 1 & 2/3/23 & Wile E Coyote & 123 Desert Station \\
   \hline
   Dynamite & 2 & & Daffy Duck & White Rock Lake \\
   \hline
   Bird Seed & 1 & & Wile E Coyote & 123 Desert Station \\
   \hline
  \end{tabular}
\caption{Orders}
\end{table}

\begin{table}[h!]
  \centering
    \begin{tabular}{ |l|l|}
      \hline
      Name & Address \\
      \hline \hline
      Wile E Coyote & 123 Desert Station \\
      \hline
      Daffy Duck & White Rock Lake \\
      \hline
    \end{tabular}
  \caption{Customers}
\end{table}

\begin{table}[h!]
  \centering
    \begin{tabular}{ |l|c|c|l|}
     \hline
     Item & Quantity & Ship Date & Customer name \\
     \hline \hline
     Anvil & 1 & 2/3/23 & Wile E Coyote \\
     \hline
     Dynamite & 2 & & Daffy Duck  \\
     \hline
     Bird Seed & 1 & & Wile E Coyote \\
     \hline
    \end{tabular}
  \caption{Orders linking to Customers}
\end{table}

We often encounter this sort of design change when we realize that attributes of one type of entity should actually belong to a new type of entity that will be associated with the first one. Often the motivation is to centralize changes to those attributes in one place. We call this operation \textit{Extract Entity}.

In a SQL database, \texttt{Customers.Name} would be a primary key and \texttt{Orders.Customer} would be a foreign key. But before \texttt{Customers.Name} can be made a primary key the two occurrences of \texttt{Wile E Coyote} must be deduplicated one way or another~\cite{dedupe}.

\textcolor{red}{this paragraph is now redundant with Related Owrk}
Given the dominance of SQL databases it is not surprising that there are many tools for performing schema change on them~\cite{RailsMigrations, liquibase}. We do find it surprising how low-level these tools are. There is support for the basic SQL DDL commands like adding and dropping tables, columns, or constraints, but no higher level operations that would help solve our problem. One is left to write custom SQL, which can become difficult if the migration needs to be performed online without bring down the database. There is a rich theory of relational database \textit{normalization} based on \textit{functional dependencies}~\cite{Molina08} but that theory has not been implemented into schema change operations in databases -- all you get is SQL. \citet{ambler06} define higher-level patterns of schema change like \textit{Split Table} that sketch SQL implementations but don't provide everything required for Extract Entity. These difficulties fairly beg for new linguistic abstractions, though programming patterns and APIs are reasonable pragmatic solutions to our challenge.

The Extract Entity schema change is straightforward in a spreadsheet. The \texttt{Customer Name} and \texttt{Customer Address} columns are copied and pasted into a new Customers table. The \textit{Remove Duplicates} command is used to deduplicate the customers. To preserve the original view for the shipping department the \texttt{Customer Address} column is redefined with a \texttt{VLOOKUP} formula to pull the address out of the customer table based on the customer's name. A spreadsheet won't maintain referential integrity between the two tables, but end-user database tools can do that~\cite{airtable, notion}.

\subsection{Challenge}
The Extract Entity challenge for live programming is to provide interactive operations in the programming environment that will perform the schema change on live code and data. The data can be in any form desired, including relations, objects, or JSON. The code can be a simple CRUD UI that also provides the shipping department its view of unshipped orders. The central problem is to coordinate the schema change with code edits/refactorings to keep the system executing live and correctly. Solutions should try to minimize:

@@

@@

>>>>

\usepackage{inputenx}
\usepackage[backend=biber]{biblatex}
%\usepackage{natbib}
\addbibresource{paper.bib}

@@

@@

\textcolor{red}{This section was an extension of the Extract Entity challenge but should be lifted up to the top-level discussion of divergence}

@@

\section{Challenge goals}
\textcolor{red}{Text from the Extract Entity challenge that maybe should be lifted up to the top-level discussion of goals}

The Extract Entity challenge for live programming is to provide interactive operations in the programming environment that will perform the schema change on live code and data. The data can be in any form desired, including relations, objects, or JSON. The code can be a simple CRUD UI that also provides the shipping department its view of unshipped orders. The central problem is to coordinate the schema change with code edits/refactorings to keep the system executing live and correctly. Solutions should try to minimize:

Some tools provide a Domain Specific Language (DSL) to describe schema evolution. Rails Migrations~\cite{RailsMigrations} embeds a DSL in Ruby to describe a range of common schema evolutions. However that falls short of the range of evolutions described by \citet{ambler06}, requiring the addition of custom Ruby or SQL code. \citet{curino08} spawned a stream of research on Database Evolution Languages (DEL) by defining a Schema Modification Operator (SMO) as ``a function that receives as input a relational schema and the underlying database, and produces as output a (modified) version of the input schema and a migrated version of the database''. SMOs can also rewrite queries to accomodate schema changes. \citet{herrmann15} defined a relationally complete DEL and then extended it into a Bidirectional Database Evolution Language (BiDEL)~\cite{herrmann17}. BiDEL appears capable of handling the basic requirements of our \textit{Extract/Absorb Entity} and \textit{Multiplicity Change} challenges, though \textit{Split/Merge Entity} is less clear. It provides schema divergence by supporting multiple schema within one database, but would need some extension to handle data divergence.

Schema evolution is also a problem for NoSQL~\cite{sadalage12} databases. While such databases are sometimes called ``schemaless'' in effect that means the schema is left implicit and tools must try to infer it~\cite{storl20, storl22}. \citet{Cambria} uses lenses~\cite{Foster2007} for bidirectional transformation of JSON. \citet{scherzinger13} define a set of operators like the relational SMOs discussed above. \citeauthor*{chillon21}~\cite{chillon21, chillon22} offer a more comprehensive set of SMOs that may be capable of handling \textit{Extract/Absorb Entity} and \textit{Multiplicity Change} but not \textit{Split/Merge Entity} nor divergence. None of the above approaches to NoSQL evolution have yet extended into updating code or rewriting queries like their SQL cousins.

\section{Extract/Absorb/Split/Merge Entity}
\textcolor{red}{Maybe called \textit{Individuation}? Maybe called \textit{Entity Evolution}?}

The Acme Corporation needs to record orders from various customers for various products. The simplest implementation is a spreadsheet with a row for each order and columns for information about the customer and product. It might look like this:

\begin{table}[h!]
  \sffamily
  \small
  % \centering
  \begin{tabular}{ |r|l|r|r|l|l|}
   \hline
   order\_id & item & quantity & ship\_date & customer\_name & customer\_address \\
   \hline \hline
   1 & Anvil & 1 & 2/3/23 & Wile E Coyote & 123 Desert Station \\
   \hline
   2 & Dynamite & 2 & & Daffy Duck & White Rock Lake \\
   \hline
   3 & Bird Seed & 1 & & Wile E Coyote & 123 Desert Station \\
   \hline
  \end{tabular}
  \caption{Orders}
\end{table}

\begin{table}[h!]
  \sffamily
  \small
  % \centering
    \begin{tabular}{ |r|l|l|}
      \hline
      customer\_id & customer\_name & customer\_address \\
      \hline \hline
      1 & Wile E Coyote & 123 Desert Station \\
      \hline
      2 & Daffy Duck & White Rock Lake \\
      \hline
    \end{tabular}
  \caption{Customers}
\end{table}

\begin{table}[h!]
  \sffamily
  \small
  % \centering
    \begin{tabular}{ |r|l|r|r|r|}
     \hline
     order\_id & item & quantity & ship\_date & customer\_id \\
     \hline \hline
     1 & Anvil & 1 & 2/3/23 & 1 \\
     \hline
     2 & Dynamite & 2 & & 2  \\
     \hline
     3 & Bird Seed & 1 & & 1 \\
     \hline
    \end{tabular}
  \caption{Orders linking to Customers}
\end{table}

We often encounter this situation when we realize that the attributes of one type of entity should actually belong to a distinct type of entity that will be associated with the first one. Often the motivation is to centralize changes to those attributes in one place. That is what it means to be an \textit{entity}: a referenceable holder of mutable attributes. We call this operation \textit{Extract Entity}.

Unique identifiers (here consecutive numbers) have been assigned to the customers and are referenced from the orders. This allows, for example, a spelling error in a customer address to be fixed in one place. It also allow a customer name or address to change without breaking the connection from all associated orders. Unique identifiers are the essence of entities, whether they are uniquely generated primary keys in a database or an abstraction of a memory address in a programming language.

The \textit{Extract Entity} operation must 1) merge duplicates, 2) assign unique identifiers, and 3) reference these identifiers from the orders. Our example took the simple approach of merging entities whose attributes are all equal. But that might not be correct in all cases. What if there was a typo in one of the addresses that made it look different from other instances of the customer? To repair that error we need to merge the falsely distinct customers into one and fix any references from orders. We call this operation \textit{Merge Entities} -- note that it is not a schema change but a data change, yet nevertheless not commonly supported in data APIs.

Acme Corp has a problem, for spreadsheets don't naturally support relationships between entities. End-user databases like Airtable~\cite{airtable} and Notion~\cite{notion} support \textit{links} between tables but do not provide operations that can do an \textit{Extract Entity}. Neither do the established database products (see Related Work). Acme will need to write code or do a big manual conversion.

It is natural to expect every schema evolution to have an inverse, for the change in requirements that triggered an evolution might change back again, and we ought to pay off the technical debt of obsolete design decisions. Research on \textit{Divergence Control}~\cite{Foster2007, herrmann17, chillon22} further suggests operations should be invertible.
The inverse of \textit{Extract Entity} is \textit{Absorb Entity}. It raises the question of how to handle \textit{one-to-many} relationships. For example should absorbing orders into customers do a relational join, returning us back to the starting point, or should it nest orders within customers, as a NoSQL database might do?

The inverse of \textit{Merge Entities} is \textit{Split Entity}, but that can be awkward. Imagine that we didn't have addresses for customers to disambiguate them with and only knew their non-unique names. Then \textit{Extract Entity} will conflate them and discard the information needed to properly invert the operation. Should this information be retained? One could object that we shouldn't have taken orders for people we can't uniquely identify yet. And even if that was necessary it means the customers really aren't distinct entities and shouldn't have been extracted. These are arguments that we shouldn't need \textit{Split Entity}, contradicting the principle that evolution operations should be invertible. It is interesting to observe that evolution in nature can also fail to invert, leaving behind vestigial features. We consider this to be an unsettled issue.

\subsection{Code Coevolution}
A simple example of how code should evolve along with the schema is a SQL query to report all pending orders in the original schema:

\begin{lstlisting}[language=SQL]
SELECT order_id, item, quantity, customer_name, customer_address
FROM Orders WHERE ship_date IS NULL;
\end{lstlisting}

The \textit{Extract Entity} operation described earlier should rewrite this query into:

\begin{lstlisting}[language=SQL]
SELECT order_id, item, quantity, customer_name, customer_address
FROM Orders
JOIN Customers ON Orders.customer_id = Customers.customer_id
WHERE ship_date IS NULL;
\end{lstlisting}

Alternatively the \textit{Extract Entity} operation could create a view in which the original query is executed.

\subsection{Goals}
A solution to this challenge should:
\begin{enumerate}
  \item Define the \textit{Extract Entity} operation.
  \item Define the \textit{Absorb Entity} operation and motivate the tradeoff between joining and nesting.
  \item Define a \textit{Merge Entities} operation or if using automatic deduplication then discuss how to handle mistakes.
  \item Define the \textit{Split Entity} operation or justify why it is not needed.
  \item If proposing a user-facing solution explain the UI affordances invoking these operations.
  \item Extend these operations to the Divergence Control scenario.
\end{enumerate}
